Visualizing Multi-way Contingency Tables

### 2005-11-17 trunk/vcd/paper/hsc.Rnw
Further improvements
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@499 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

<<preliminaries,echo=FALSE,results=hide>>=
set.seed(1071)
library(vcd)
library(scatterplot3d)
library(xtable)
data(Arthritis)
data(Titanic)
data(PreSex)
data(UCBAdmissions)
hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
dimnames = list("Visit frequency" = c("Regular","Less than monthly","Never"), 
"Length of stay" = c("2-9","10-19","20+"))))
art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")
@

@@

This chapter will give an overview of the state of the art 
of mosaic, sieve, and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. Sect.~\ref{sec:concl} concludes the chapter.

@@

\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
significantly different in the ``less than monthly'' and ``never''
strata. From the row-standardized table (see Tab~.\ref{tab:rowhospital}), 
it seems indeed that LOS were homogeneous with respect to these two visit frequency strata.

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a squared mosaic. The tiles are
obtained by recursive partitioning splits of a square.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is inversed.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
"dotted", col = "dark gray"))
labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Sieve plot for the \data{hospital} data.}
\label{fig:sievefig}
\end{center}
\end{figure}

@@

\section{Using colors for Residual-based Shadings}
\label{sec:extensions}

@@

The plots introduced in the previous section are basically composed of
tiles whose areas represent characteristics derived from the
contingency tables---observed and expected frequencies in the case of
mosaic and sieve plots, or residuals as visualized by association
plots. When using color for these tiles, it is imperative
to choose the right color palettes, derived from suitable color
spaces. Apart from esthetic considerations, wrongly chosen colors
might serously affect the analysis: For example,

\begin{itemize}
\item Using high-chroma colors for large
  areas tend to produce after-image effects which can be
  distracting \cite{vcd:Ihaka:2003}.
\item Lighter colors tend to make areas look larger than darker
  colors, therefore using colors with unequal luminance values  make it
  difficult to compare area sizes \cite{vcd:Cleveland+McGill:1983}.
\item Color palettes derived from
  non-uniform color spaces may contain unbalanced colors with respect to
  their colorfulness or lightness. When tiles are shaded using such a
  palette, some of them might appear more important than others in an
  uncontrolled way.
\end{itemize}

% gamma correction for displays.

\noindent Implementations are often based on so-called
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
color scheme). Both spaces are rather similar transformations of RGB (Red-Green-Blue) space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions: the hue (dominant wavelength in the spectrum), 
the saturation (`colorfulness' or `pureness'), and 
the value (`brightness', amount of gray). These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red over a neutral white
to a saturated blue. Although the palette should be balanced with
respect to colorfulness and lightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

The use of colors that are more `in harmony' go back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
Chroma loosely corresponds to colorfulness and luminance to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and dividing the hue range in equally-sized intervals. 
HCL colors with fixed chroma and luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} that are
suited, e.g., for visualizing residuals as we will see in the next
section. In principle, for each of the two bipolar hues of the
diverging palette, we could hold constant 
the luminance and only vary the chroma to get the different colors. 
However, it turns out that the resulting colors 
are too similar to be used for discriminating shaded
objects. Better contrasting effects can be achieved by \emph{simultaneously}
varying chroma and luminance, for example by choosing two points in
the chroma/luminance plane for a given hue and interpolating
linearly. For each half of the diverging palette, 
one of these points should be a ``neutral'' color,
and the other correspond to the ending part of the scale. 
The lower part of Fig.~\ref{fig:diverging} shows a diverging
palette using HCL colors. In contrast to the HSV counterpart, matching
colors are balanced to the same gray, and thus get the same perceptual
``weight''.

\subsection{Using Colors for Residuals}
\label{sec:shadings}

All plots introduced in Section~\ref{sec:twoway} can be improved using colors,
either by adding additional information, or by redundantly coding information
already visualized by the `raw' plot to support our perceptual system. %'

First, we consider the sieve plots. The density 
of the grid in the raw version implicitly gives us an 
idea of the residuals' %'
size, but since the plot does not include the density corresponding to zero
residuals (the null model) for comparison, we 
cannot easily assess whether there are more, of less counts in a cell 
than expected under the null hypothesis. It would help if we knew the 
sign of the residuals. 
Using color, we can add this information, for example using blue for
positive, red for negative, and gray for zero residuals. Figure~\ref{fig:colorsieve} 
demonstrates the effect of applying such a color shading to a sieve plot for the 
\data{hospital} data. Color clearly emphasizes the linear association
pattern in this data: Patients with a LOS from 2--9 years are
associated with regular visits, whereas patients with a LOS from
10--19 and more than 19 years are associated with a visit frequency
of less than monthly and no visits.

@@

Mosaic plots in their initial version are monochrome
displays. Friendly \cite{vcd:Friendly:1994} introduced a
residual-based shading of the tiles to additionally visualize the
residuals from a given independence model fitted to the table. 
The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) have no color. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red, respectively. 
The heuristic for choosing the cut offs $2$ and $4$ is that the Pearson residuals
are asymptotically standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of the
Friendly shading is not to visualize significance but the \emph{pattern} of
deviation from independence \cite{vcd:Friendly:2000}. 
In addition to the shading of the rectangles themselves, the Friendly shading also
encompasses a choice of line type and line color of the rectangle borders
with similar ideas as described above, useful when no color can be used.

In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
data, this time using a Friendly-like color coding (with HCL
instead of HSV colors, and no line type coding).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<assoccolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
assoc(t(hospital), split = TRUE, shade = TRUE, keep = TRUE)
@ 
\caption{Association plot with Friendly-like color coding of the residuals.}
\label{fig:colorassoc}
\end{center}
\end{figure}

@@

Figure~\ref{fig:Friendly} includes the $p$ value of a $\chi^2$ test
of independence, frequently used to assess the significance of the
hypothesis of independence (or homogeneity for stratified data) in
two-way tables. The test statistic is just the sum of the squared
Pearson residuals

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using darker (`uninteresting') colors for non-significant results by
restricting the luminance range used for interpolating the colors.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Tab.~\ref{tab:arthritis}):

@@

\noindent Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $99\%$ level ($p=0.0035$), but
since all residuals are in the $[-1.72,1.87]$ interval, the tiles
remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the
sum of squares:

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
The conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Fig.~\ref{fig:arthritis2}, we again
visualize the \data{arthritis} data, this
time using the maximum test statistic. The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, shade = TRUE)
@
\caption{Mosaic plot for the \data{Arthritis} data, using the $\chi^2$ test 
and fixed cut-off points for the shading.}
\label{fig:arthritis}
\end{center}
\end{figure}

@@

>>>>

%% Hyperref
\hypersetup{%
  pdftitle = {Residual-based Shadings for Visualizing (Conditional) Independence},
  pdfsubject = {Residual-based Shadings},
  pdfkeywords = {association plots, conditional inference, contingency tables, HCL colors, HSV colors, mosaic plots},
  pdfauthor = {Achim Zeileis, David Meyer, Kurt Hornik},
  %% change colorlinks to false for pretty printing
  colorlinks = {false},
  linkcolor = {Blue},
  citecolor = {Blue},
  urlcolor = {Red},
  hyperindex = {true},
  linktocpage = {true},
}

<<preliminaries,echo=FALSE,results=hide>>=
myseed <- 4711
set.seed(myseed)
library(vcd)
library(scatterplot3d)
library(xtable)
data(Arthritis)
data(UCBAdmissions)
hospital <- as.table(matrix(c(43,16,3,6,11,10,9,18,16), c(3,3), byrow = TRUE, 
dimnames = list("Visit frequency" = c("Regular","Less than monthly","Never"), 
"Length of stay" = c("2-9","10-19","20+"))))
art <- xtabs(~Treatment + Improved, data = Arthritis, subset = Sex == "Female")

set.seed(myseed)
art_max <- coindep_test(art, n = 5000)
set.seed(myseed)
art_chisq <- coindep_test(art, n = 5000, indepfun = function(x) sum(x^2))

set.seed(myseed)
hos_chisq <- coindep_test(hospital, n = 5000, indepfun = function(x) sum(x^2))

set.seed(myseed)
hec_max <- coindep_test(HairEyeColor, "Sex", n = 5000)

@

@@

This chapter will give an overview of the state of the art 
of mosaic, sieve, and association plots, both for exploratory visualization and
model-based analysis. Exploratory techniques will include specialized displays for
the bivariate case, as well as pairs plot-like displays for
higher-dimensional tables. As for the model-based tools, particular
emphasis will be given to methods suitable for the visualization of 
conditional independence tests (including permutation tests), as well
as for the visualization of particular GLMs (such as log-linear models). 
In Sect.~\ref{sec:twoway}, we start with the
simple bivariate case.  Section~\ref{sec:extensions} explains how the use of color
in residual-based shadings can support data exploration, and
even promotes the methods to diagnostic and model-based tools
by visualizing test statistics and residuals of independence models.
In Sect.~\ref{sec:multiway}, we show how the
basically bivariate methods straightforwardly extend to the multivariate case by using
`flat' representations of the multi-way tables. In this section, we
also introduce specialized displays for conditional independence
structures. Section~\ref{sec:concl} concludes the chapter.

@@

\noindent In addition, \cite{vcd:haberman:1974} notes that this pattern is not
significantly different in the ``less than monthly'' and ``never''
strata. From the row-standardized table (see Tab.~\ref{tab:rowhospital}), 
it seems indeed that LOS were homogeneous with respect to these two visit frequency strata.

@@

Mosaic displays have been introduced by 
\cite{vcd:Hartigan+Kleiner:1981,vcd:Hartigan+Kleiner:1984} and
extended, e.g., by \cite{vcd:Friendly:1994,vcd:Friendly:1999,vcd:Friendly:2000}. 
They visualize the observed values of a contingency table by
area-proportional tiles, arranged in a squared mosaic. The tiles are
obtained by recursive partitioning splits of a square.
In the following, we describe the main idea of mosaic plots, 
chapter XXX in this book \emph{\bf (Reference to other contributions)} 
provides more detailed information. Consider our
example of the hospital data from above. Step 1 consists of splitting a square
according to the marginals of one of the variables. To be consistent with the textual
representation, we choose LOS with vertical splits (see Fig.~\ref{fig:mosaic1fig}). 
The result is similar to a bar plot where not the height, but the width is
adapted to visualize the counts for each level, which is also 
called a \emph{spine plot} \cite{vcd:hummel:1996}. From this plot, we see that 
the number of patients 
decreases with the length of stay. Step 2 now is to add further splits in the
other direction, i.e., horizontal splits, for the second
variable. This means that each vertical bar is split according to the
marginals of the second variable, \emph{given} the first variable (see
Fig.~\ref{fig:mosaic2fig}). The resulting plot visualizes the 
contingency table where each cell has a size proportional to the
corresponding table entry.
We can still see the marginal distribution of LOS, %'
and additionally, the visit frequency \emph{given} each category of LOS. %'
If the two variables were independent, the grid would be regular. Clearly, compared
to a length of stay of 10--19 years, more patients get regular visits for stays from 2--9 
years, and conversely, less patients get regular visits for stays for more than 19 years.
For patients that get no visits, the pattern is reversed.

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<sievefig,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
sieve(t(hospital), split = TRUE, pop = FALSE, gp = gpar(lty =
"dotted", col = "black"))
labeling_cells(text = t(hospital), clip = FALSE, gp = gpar(fontface = 2, fontsize = 15))(t(hospital))
@ 
\caption{Sieve plot for the \data{hospital} data.}
\label{fig:sievefig}
\end{center}
\end{figure}

@@

\section{Using Colors for Residual-based Shadings}
\label{sec:extensions}

@@

The plots introduced in the previous section are basically composed of
tiles whose areas represent characteristics derived from the
contingency tables---observed and expected frequencies in the case of
mosaic and sieve plots, or residuals as visualized by association
plots. When using color for these tiles, it is imperative
to choose the right color palettes, derived from suitable color
spaces. Apart from esthetic considerations, wrongly chosen colors
might serously affect the analysis. For example,

\begin{itemize}
\item Using high-chroma colors for large
  areas tend to produce after-image effects which can be
  distracting \cite{vcd:Ihaka:2003}.
\item Lighter colors tend to make areas look larger than darker
  colors, therefore using colors with unequal luminance values  makes it
  difficult to compare area sizes \cite{vcd:Cleveland+McGill:1983}.
\item Color palettes derived from
  non-uniform color spaces may contain unbalanced colors with respect to
  their colorfulness or brightness. When tiles are shaded using such a
  palette, some of them might appear more important than others in an
  uncontrolled way.
\end{itemize}

\noindent 
Due to the three-dimensional nature of human color
perception \cite{vcd:Mollon:1995}, 
it has been common to specify colors using the three primaries
red, green, and blue (RGB colors), especially for computer devices. 
The appearance of the on-screen colors is affected by the device
characteristics: For example, the intensity $I$ of a primary 
on a particular device follows the rule $I=L^\gamma$, with $L$
the value for the primary color, and $\gamma$ device-dependent 
(but typically close to 2.2). Therefore, a first caveat is that if colors are to
appear identical on different devices, their gamma characteristics
have to be taken into account.

Now, software implementations are often based on the more convenient
Hue-Saturation-Value (HSV) colors (or the comparable Hue-Luminance-Saturation 
color scheme). Both spaces are rather similar transformations of the RGB space
\cite{vcd:Brewer:1999,vcd:Poynton:2000} and are very common implementations of
colors in many computer packages \cite{vcd:Moretti+Lyons:2002}. Each color in this space is
represented by three dimensions: the hue (dominant wavelength in the spectrum), 
the saturation (`colorfulness' or `pureness'), and 
the value (`brightness', amount of gray). These intuitive dimensions
make HSV colors easyier to specifiy than, e.g, RGB colors. 
But HSV colors have several disadvantages.
Most importantly, HSV colors are not perceptually
uniform because the three HSV dimensions map only poorly to the three
perceptual dimensions of the human visual system  \cite{vcd:Brewer:1999,vcd:Ihaka:2003}.
One important issue here is that HSV dimensions are confounded, e.g., saturation
is not uniform across different hues.
As an example, see Fig.~\ref{fig:colors} (left) 
showing a qualitative color palette in the HSV
space: although saturation and value are fixed, the fully saturated blue is
perceived much darker than the fully saturated red or green,
making it difficult to judge the size of shaded areas.
Furthermore, the flashy fully saturated HSV colors are hard 
to look at for a longer time. For similar reasons, it is equally difficult
to derive acceptable diverging palettes from the HSV space, i.e.,
bipolar scales containing colors ranging between two very distinct colors.
The upper part of Fig.~\ref{fig:diverging} shows a diverging
palette in the HSV space with colors ranging from a saturated red over a neutral white
to a saturated blue. Although the palette should be balanced with
respect to colorfulness and brightness, the red colors are perceived
to be more more intense and flashy than the corresponding blue
colors.

@@

The use of colors that are more `in harmony' go back to Munsell 
\cite{vcd:Munsell:1905} %'
who introduced a notation for balanced colors. Based on those,
tools producing better palettes for specific tasks have been developed 
\cite{vcd:Harrower+Brewer:2003}. Other perceptually-based color
spaces, especially suited for computer displays, are the CIELAB and CIELUV
spaces \cite{vcd:CIE:2004} from which qualitative palettes for
statistical graphics have been derived \cite{vcd:Ihaka:2003}. A
transformation of the CIELUV space leads to the HCL (Hue-Chroma-Luminance) space.
Chroma loosely corresponds to colorfulness and luminance to
brightness, but in contrast to HSV colors, chroma is an
\emph{absolute} measure valid for all hues, and luminance can be
varied independently from the other two dimensions.
Qualitative color palettes can easily be obtained by holding constant
chroma and luminance, and dividing the hue range in equally-sized intervals. 
HCL colors with fixed chroma and luminance 
are always balanced towards the same grey 
and thus do not have the problem of varying saturations as the HSV
colors (see Fig.~\ref{fig:colors}, right). 
Similarly, diverging color palettes can be derived
\cite{vcd:Zeileis+Meyer+Hornik:2005} that are
suited, e.g., for visualizing residuals as we will see in the next
section. In principle, for each of the two bipolar hues of the
diverging palette, we could hold constant 
the luminance and only vary the chroma to get the different colors. 
However, it turns out that the resulting colors 
are too similar to be used for discriminating shaded
objects. Better contrasting can be achieved by \emph{simultaneously}
varying chroma and luminance, for example by choosing two points in
the chroma/luminance plane for a given hue and interpolating
linearly. For each half of the diverging palette, 
one of these points should be a ``neutral'' color,
and the other correspond to the ending part of the scale. 
The lower part of Fig.~\ref{fig:diverging} shows a diverging
palette using HCL colors. In contrast to the HSV counterpart, matching
colors (light red/blue and dark red/blue) 
are balanced to the same gray, and thus get the same perceptual ``weight''.

\subsection{Highlighting and Color-based Shadings}
\label{sec:shadings}

All plots introduced in Sect.~\ref{sec:twoway} are composed of empty
tiles. It seems intuitive to used filled tiles to highlight information
of interest. Consider again the hospital example: in Fig.~\ref{fig:highlight},
we mark tiles for patients never or seldom visited to visualize their proportion in
the LOS strata. For optical clarity, we set the
spacing between the visit frequency tiles to 0. Clearly, the
proportion of these patients increases with LOS. In fact, such a
``stacked'' plot can also be interpreted as spine plot with highlighting
\cite{vcd:hummel:1996}, particularly useful in analyzing causal relationships in
categorical data with a binary dependent variable. More variations on
this theme, such as doubledecker plots, are treated in chapter XXX
(\textbf{REFERENCE TO OTHER CHAPTER(S)}).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<highlight,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), 
       mar = c(left = 3.5), 
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))),
       split = TRUE, spacing = spacing_highlighting, 
       gp = gpar(fill = c("gray","black","black")))
@ 
\caption{Spine plot with highlighting for the hospital data.}
\label{fig:highlight}
\end{center}
\end{figure}

Using colors, even more complementary information can be visualized,
either by adding additional information, or by redundantly coding information
already visualized by the `raw' plot to support our perceptual system. %'
First, we consider the sieve plots. The density 
of the grid in the raw version implicitly gives us an 
idea of the residuals' %'
size, but since the plot does not include the density corresponding to zero
residuals (the null model) for comparison, we 
cannot easily assess whether there are more, of less counts in a cell 
than expected under the null hypothesis. It would help if we knew the 
sign of the residuals. 
Using color, we can add this information, for example using blue for
positive, red for negative, and gray for zero residuals. Figure~\ref{fig:colorsieve} 
demonstrates the effect of applying such a color shading to a sieve plot for the 
\data{hospital} data. Color clearly emphasizes the linear association
pattern in this data: Patients with a LOS from 2--9 years are
associated with regular visits, whereas patients with a LOS from
10--19 and more than 19 years are associated with a visit frequency
of less than monthly and no visits.

@@

Mosaic plots in their initial version are monochrome
displays. Friendly \cite{vcd:Friendly:1994} introduced a
residual-based shading of the tiles to additionally visualize the
residuals from a given independence model fitted to the table. 
The idea is to use a color coding for the mosaic tiles that visualizes the
sign and absolute size of each residual $r_{ij}$: Cells corresponding to small
residuals ($|r_{ij}| < 2$) have no color. Cells with medium sized residuals
($2 \le |r_{ij}| < 4$) are shaded light blue and light red for positive and negative
residuals, respectively. Cells with large residuals ($|r_{ij}| \ge 4$) are shaded
with a fully saturated blue and red, respectively. 
The heuristic for choosing the cut-offs $2$ and $4$ is that the Pearson residuals
are asymptotically standard normal which implies that the highlighted cells are those with
residuals \emph{individually} significant at approximately the $\alpha
= 0.05$ and $\alpha = 0.0001$ levels. However, the main purpose of the
Friendly shading is not to visualize significance but the \emph{pattern} of
deviation from independence \cite{vcd:Friendly:2000}. 
In addition to the shading of the rectangles themselves, the Friendly shading also
encompasses a choice of line type and line color of the rectangle borders
with similar ideas as described above, useful when no color can be used.

In Fig.~\ref{fig:Friendly}, we again show the mosaic for the \data{hospital}
data, this time using a Friendly-like color shading (with HCL
instead of HSV colors, and no line type coding).

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<friendly,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(t(hospital), split = TRUE, shade = TRUE, 
       mar = c(left = 3.5), 
       gp_args = list(p.value = hos_chisq$p.value),
       labeling_args = list(offset_labels = c(left = 0.5), 
         offset_varnames = c(left = 1, top = 0.5), set_labels = 
         list("Visit frequency" = c("Regular","Less than\nmonthly","Never"))))
@ 
\caption{Mosaic display with Friendly-like color coding of the residuals.}
\label{fig:Friendly}
\end{center}
\end{figure}

@@

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}
\begin{center}
<<assoccolor,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
assoc(t(hospital), split = TRUE, shade = TRUE, keep = TRUE,
      gp_args = list(p.value = hos_chisq$p.value))
@ 
\caption{Association plot with Friendly-like color coding of the residuals.}
\label{fig:colorassoc}
\end{center}
\end{figure}

@@

Figures~\ref{fig:Friendly} and \ref{fig:colorassoc} 
include the (same) $p$ value of a $\chi^2$ test
of independence, frequently used to assess the significance of the
hypothesis of independence (or homogeneity for stratified data) in
two-way tables. The test statistic is just the sum of the squared
Pearson residuals

@@

\noindent known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
An important reason for using the unconditional limiting distribution
for the $X^2$ statistic from Equ.~\ref{eq:ChiSquare} was the closed form result
for the distribution. Recently, with the improving perfomance of computers,
conditional inference (or permutation tests)---carried out either by simulation
or by computation of the (asymptotic) permutation distribution---have been receiving
increasing attention \cite{vcd:Ernst:2004,vcd:Pesarin:2001,vcd:Strasser+Weber:1999}. 
For testing the independence hypothesis from Equation~\ref{eq:H0},
using a permutation test is particularly intuitive due to the 
permutation invariance (given row and column sums) of this problem.
Consequently, all results in this paper are based on conditional inference performed
by simulating the permutation distribution of test statistics of type
$\lambda([r_{ij}])$.

Since the HCL space is three-dimensional and we only
used two `degrees of freedom' so far for coding information %'
(hue for the sign and a linear combination of chroma and luminance 
for the size of the residuals), we can add a third information to the plot.
For example, we can visualize the significance of some specified test statistic
(e.g., the $\chi^2$ test statistic) 
using darker (`uninteresting') colors for non-significant
results. These can again be derived using the interpolation procedure
described in the Sect.~\ref{sec:colors} by
restricting the luminance to a shorter range, ending with a ``darker'' value.

@@

The heuristic for choosing the cut-off points 
in the Friendly shading may lead to wrong conclusions: especially in
large tables, the test of
independence may not be significant even though some of the residuals are
``large''. On the the other hand, the test might be significant even
though the residuals are ``small''. In fact, the cut-off points are really
data-dependent. Consider the case of the \data{arthritis} data 
\cite{vcd:Koch+Edwards:1988}, resulting from a double-blind clinical trial
investigating a new treatment for rheumatoid arthritis, stratified by
gender (see Tab.~\ref{tab:arthritis} for the female patients):

@@

\noindent Figure~\ref{fig:arthritis} visualizes the results for the female
patients, again by means of a mosaic display. Clearly, the hypothesis
of independence is rejected by the $\chi^2$ test even on a $99\%$ level 
($p=\Sexpr{round(art_chisq$p.value,4)}$), but
since all residuals are in the 
$[\Sexpr{round(min(chisq.test(art)$residuals),4)},\Sexpr{round(max(chisq.test(art)$residuals),4)}]$ %$
interval, the tiles remain uncolored. A solution to this issue is to use a different
test statistic, for example the maximum of the absolute values of the
Pearson residuals \cite{vcd:Meyer+Zeileis+Hornik:2003} instead of the sum of squares:

@@

\noindent Given a critical value $c_\alpha$ for this test statistic, all residuals whose
absolute value exceeds $c_\alpha$ violate the hypothesis of indendence at level
$\alpha$ \cite[ch. 7]{vcd:Mazanec+Strasser:2000}. Thus, the interesting cells giving
evidence for the rejection of the independence hypothesis can easily be identified.
As explained above, 
the conditional distribution of this test statistic under the null can be obtained
by simulation, sampling tables with the same row and column sums $n_{i+}$ and $n_{+j}$
using, e.g., the Patefield algorithm \cite{vcd:Patefield:1981} 
and computing the maximum statistic for each of these tables.
In Fig.~\ref{fig:arthritis2}, we again
visualize the \data{arthritis} data, this
time using the maximum test statistic. The shading of the
tiles now clearly shows that the treatment is effective: significantly more patients
in the treatment group exhibit marked improvement than would be
expected under independence.

\setkeys{Gin}{width=0.7\textwidth}
\begin{figure}[p]
\begin{center}
<<Arthritis,fig=TRUE,echo=FALSE,height=6,width=7,results=hide>>=
mosaic(art, shade = TRUE, gp_args = list(p.value = art_chisq$p.value))
@
\caption{Mosaic plot for the \data{Arthritis} data, using the $\chi^2$ test 
and fixed cut-off points for the shading.}
\label{fig:arthritis}
\end{center}
\end{figure}

@@

Especially for the visualization of areas, 
use perceptually uniform colors and color palettes, derived from,
e.g., the HCL color space. Shadings can be used to add 
information to the basic plots and to support the
analysis. Highlighting of tiles can support the analysis of causal relationships.
Residual-based shadings can be used for model diagnostics, e.g.,
by using the hue to code the sign of a residual, and a linear
combination of chroma and luminance for the size. In addition, the
significance of test statistics can be visualized through overall 
darker color palettes. Using the maximum instead of the $\chi^2$ test
statistic to aggregate the residuals allows precise diagnostics when
the null hypothesis of independence is rejected.
